---
title: "Getting Started with divaodv"
author: "Ilia Maidanik"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: false
    fig-width: 10
    fig-height: 5
    fig-dpi: 150
vignette: >
  %\VignetteIndexEntry{Getting Started with divaodv}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
execute:
  eval: false
  echo: true
---

```{r}
#| label: setup
#| include: false
library(divaodv)
```

## Overview

**divaodv** creates Ocean Data View (ODV) style depth × time section plots from
scattered oceanographic profile data. It uses DIVAnd (Data-Interpolating
Variational Analysis) via Julia to interpolate observations onto a regular grid,
then renders the result with ggplot2.

## Prerequisites

Before using divaodv, you need Julia and DIVAnd.jl installed. Run the built-in
check:

```{r}
#| label: setup-check
diva_setup_check()
```

## Minimal example

The package ships with a small temperature dataset from the NMP Gulf of Aqaba
Station A monitoring programme:

```{r}
#| label: load-data
df <- read.csv(system.file("extdata", "nmp_temp_200m.csv", package = "divaodv"))
df$Date <- as.Date(df$Date)

str(df)
```

The data frame needs three columns: `Date` (Date class), `depth` (numeric, metres),
and at least one variable column.

## Your first ODV plot

```{r}
#| label: basic-plot
#| fig-cap: "Temperature section — Gulf of Aqaba, upper 200 m."
p <- diva_plot_odv(
  df         = df,
  var        = "Temp",
  time_corr  = 20,
  depth_corr = 15,
  epsilon2   = 0.01,
  max_depth  = 200,
  palette    = "odv"
)
print(p)
```

That's it — one function call from raw profiles to a publication-quality section
plot.

## Understanding the parameters

### Correlation lengths

The two most important parameters control how far DIVAnd smooths in each
dimension:

- **`time_corr`** — temporal correlation length in **days**. Larger values
  produce smoother fields in time. For physical variables like temperature
  (which change slowly), 15–30 days works well. For patchy biological variables
  like chlorophyll, 5–10 days preserves mesoscale structure.

- **`depth_corr`** — depth correlation length in **metres**. Controls vertical
  smoothing. Typically 10–20 m for the upper water column.

### Signal-to-noise ratio (ε²)

`epsilon2` controls the balance between fitting observations and producing a
smooth field. Smaller values (0.001–0.01) trust the data more; larger values
(0.1–1.0) produce smoother fields. For quality-controlled monitoring data,
0.01 is a good starting point.

### Log transforms

Many biological and chemical variables (chlorophyll, nutrients, cell counts)
have right-skewed distributions. Setting `transform = "log"` applies a log
transform before interpolation and back-transforms afterward:

```{r}
#| label: log-transform
#| eval: false
diva_plot_odv(df, "Chl_a_ug_L",
              time_corr = 7, depth_corr = 10, epsilon2 = 0.02,
              transform = "log")
```

## Grid resolution

Two parameters control the interpolation grid density:

```{r}
#| label: coarse-grid
#| fig-cap: "Coarse grid (weekly × 2 m) — fast preview."
diva_plot_odv(df, "Temp", time_corr = 20, depth_corr = 15,
              time_resolution = 52, depth_resolution = 2)
```

```{r}
#| label: fine-grid
#| fig-cap: "Fine grid (daily × 1 m) — publication quality."
diva_plot_odv(df, "Temp", time_corr = 20, depth_corr = 15,
              time_resolution = 365, depth_resolution = 1)
```

Start coarse for parameter tuning, then increase resolution for the final figure.

## Customising the plot

Since `diva_plot_odv()` returns a ggplot2 object, you can add layers:

```{r}
#| label: custom-plot
#| fig-cap: "Customised temperature section."
library(ggplot2)

p <- diva_plot_odv(df, "Temp", time_corr = 20, depth_corr = 15,
                   contour_binwidth = 2, label_binwidth = 2,
                   title = "Sea Temperature — Gulf of Aqaba",
                   fill_label = "T (°C)")

p + theme(text = element_text(family = "serif"))
```

## Colour palettes

Three built-in options:

```{r}
#| label: palettes
#| eval: false
# Classic ODV rainbow (default)
diva_plot_odv(df, "Temp", time_corr = 20, depth_corr = 15, palette = "odv")

# Viridis (perceptually uniform, colourblind-safe)
diva_plot_odv(df, "Temp", time_corr = 20, depth_corr = 15, palette = "viridis")

# Custom hex vector
my_pal <- c("#2166ac", "#67a9cf", "#d1e5f0", "#fddbc7", "#ef8a62", "#b2182b")
diva_plot_odv(df, "Temp", time_corr = 20, depth_corr = 15, palette = my_pal)
```

## Extracting the grid

Set `return_data = TRUE` to get the interpolated grid as a tibble instead of a
plot — useful for further analysis or custom visualisation:

```{r}
#| label: return-data
grid <- diva_plot_odv(df, "Temp", time_corr = 20, depth_corr = 15,
                      return_data = TRUE)
head(grid)
```

## Multi-variable workflows

For projects with many variables, use `diva_variable_config()` to create
reusable parameter sets:

```{r}
#| label: multi-var
#| eval: false
# Load the 12-variable NMP defaults
cfg <- nmp_default_config()

# Plot each variable
for (v in names(cfg)) {
  cv <- cfg[[v]]
  p <- diva_plot_odv(
    df         = my_data,
    var        = cv$var,
    time_corr  = cv$time_corr,
    depth_corr = cv$depth_corr,
    epsilon2   = cv$epsilon2,
    transform  = cv$transform
  )
  print(p)
}
```

## How the metric tensor works

A common pitfall when using DIVAnd is setting the metric tensor (`pmn`) to
all ones. This tells DIVAnd that each grid cell is 1 × 1 in physical units,
regardless of actual spacing — which breaks the correlation length
interpretation and produces vertical-stripe artefacts.

divaodv computes `pmn` correctly from the actual grid spacing:

```
dz = max_depth / (n_depth - 1)     # metres per cell
dt = max_time  / (n_time  - 1)     # years per cell
pm = 1 / dz                        # points per metre
pn = 1 / dt                        # points per year
```

This means correlation lengths are always honoured regardless of grid
resolution — you can change `time_resolution` from 365 to 52 without
affecting the physical smoothing behaviour.
