---
title: "Getting Started with divaodv"
author: "Ilia Maidanik"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: false
    fig-width: 10
    fig-height: 5
    fig-dpi: 150
vignette: >
  %\VignetteIndexEntry{Getting Started with divaodv}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
execute:
  eval: true
  echo: true
  cache: true
---

```{r}
#| label: setup
#| include: false
library(divaodv)
```

## Overview

**divaodv** creates Ocean Data View (ODV) style depth × time section plots from
scattered oceanographic profile data. It uses DIVAnd (Data-Interpolating
Variational Analysis) via Julia to interpolate observations onto a regular grid,
then renders the result with ggplot2.

## Prerequisites

Before using divaodv, you need Julia and DIVAnd.jl installed. Run the built-in
check:

```{r}
#| label: setup-check
diva_setup_check()
```

## Minimal example

The package ships with a small temperature dataset from the NMP Gulf of Aqaba
Station A monitoring programme:

```{r}
#| label: load-data
df <- read.csv(system.file("extdata", "nmp_temp_200m.csv", package = "divaodv"))
df$Date <- as.Date(df$Date)

str(df)
```

The data frame needs three columns: `Date` (Date class), `depth` (numeric, metres),
and at least one variable column.

## Your first ODV plot

```{r}
#| label: basic-plot
#| fig-cap: "Temperature section — Gulf of Aqaba, upper 200 m."
p <- diva_plot_odv(
  df         = df,
  var        = "Temp",
  time_corr  = 20,
  depth_corr = 15,
  epsilon2   = 0.01,
  max_depth  = 200,
  palette    = "odv"
)
print(p)
```

That's it — one function call from raw profiles to a publication-quality section
plot.

## Understanding the parameters

### Correlation lengths

The two most important parameters control how far DIVAnd smooths in each
dimension:

- **`time_corr`** — temporal correlation length in **days**. Larger values
  produce smoother fields in time. For physical variables like temperature
  (which change slowly), 15–30 days works well. For patchy biological variables
  like chlorophyll, 5–10 days preserves mesoscale structure.

- **`depth_corr`** — depth correlation length in **metres**. Controls vertical
  smoothing. Typically 10–20 m for the upper water column.

### Signal-to-noise ratio (ε²)

`epsilon2` controls the balance between fitting observations and producing a
smooth field. Smaller values (0.001–0.01) trust the data more; larger values
(0.1–1.0) produce smoother fields. For quality-controlled monitoring data,
0.01 is a good starting point.

### Log transforms

Many biological and chemical variables (chlorophyll, nutrients, cell counts)
have right-skewed distributions. Setting `transform = "log"` applies a log
transform before interpolation and back-transforms afterward:

```{r}
#| label: log-transform
#| eval: false
diva_plot_odv(df, "Chl_a_ug_L",
              time_corr = 7, depth_corr = 10, epsilon2 = 0.02,
              transform = "log")
```

## Grid resolution

Two parameters control the interpolation grid density:

```{r}
#| label: coarse-grid
#| fig-cap: "Coarse grid (weekly × 2 m) — fast preview."
diva_plot_odv(df, "Temp", time_corr = 20, depth_corr = 15,
              time_resolution = 52, depth_resolution = 2)
```

```{r}
#| label: fine-grid
#| fig-cap: "Fine grid (daily × 1 m) — publication quality."
# Largest grid in this vignette; chunk caching keeps re-renders cheap.
diva_plot_odv(df, "Temp", time_corr = 20, depth_corr = 15,
              time_resolution = 365, depth_resolution = 1)
```

Start coarse for parameter tuning, then increase resolution for the final figure.

## Customising the plot

Since `diva_plot_odv()` returns a ggplot2 object, you can add layers:

```{r}
#| label: custom-plot
#| fig-cap: "Customised temperature section."
library(ggplot2)

p <- diva_plot_odv(df, "Temp", time_corr = 20, depth_corr = 15,
                   contour_binwidth = 2, label_binwidth = 2,
                   title = "Sea Temperature — Gulf of Aqaba",
                   fill_label = "T (°C)")

p + theme(text = element_text(family = "serif"))
```

## Colour palettes

Three built-in options:

```{r}
#| label: palettes
#| eval: false
# Classic ODV rainbow (default)
diva_plot_odv(df, "Temp", time_corr = 20, depth_corr = 15, palette = "odv")

# Viridis (perceptually uniform, colourblind-safe)
diva_plot_odv(df, "Temp", time_corr = 20, depth_corr = 15, palette = "viridis")

# Custom hex vector
my_pal <- c("#2166ac", "#67a9cf", "#d1e5f0", "#fddbc7", "#ef8a62", "#b2182b")
diva_plot_odv(df, "Temp", time_corr = 20, depth_corr = 15, palette = my_pal)
```

## ODV-style colour mapping {#sec-colour-mapping}

A linear colour scale wastes most of the palette when a variable is strongly
skewed — nutrients, chlorophyll and cell counts typically are. Ocean Data
View solves this with two sliders on its Color Mapping tab, and
`scale_fill_odv()` reproduces their effect:

- **`median`** — the position within the colour range that receives the
  middle colour. Below 0.5 gives more of the palette to high values; above
  0.5, to low values.
- **`nonlinearity`** — how sharply resolution concentrates around that
  median, at the expense of the range extremes. 0 is linear.

```{r}
#| label: colour-mapping-params
cm_median <- 0.35   # where in the range the middle colour sits
cm_nonlin <- 0.6    # how hard to concentrate resolution there
```

Interpolate once:

```{r}
#| label: fig-colour-linear
#| fig-cap: "Default linear colour mapping. Palette steps are evenly spaced across the value range."
p <- diva_plot_odv(df, "Temp", time_corr = 20, depth_corr = 15,
                   zlim = c(18, 28), fill_label = "T (°C)")
p
```

Then recolour it:

```{r}
#| label: fig-colour-odv
#| fig-cap: !expr sprintf("ODV-style mapping with median = %s and nonlinearity = %s. Colour resolution concentrates around the shifted midpoint.", cm_median, cm_nonlin)
p + scale_fill_odv(median = cm_median, nonlinearity = cm_nonlin)
```

Comparing @fig-colour-linear with @fig-colour-odv: both show the *same*
interpolated field over the same limits, and both colourbars carry the same
evenly spaced, round-numbered ticks. What changes is where the palette spends
its resolution. Setting `median` to `r cm_median` moves the middle colour
below the centre of the range, and `nonlinearity` at `r cm_nonlin` steepens
the transfer curve there, so gradients near the thermocline separate into
distinct bands instead of compressing into one or two.

The ticks stay linear in @fig-colour-odv. Only the colour stops move, which is
what keeps the legend readable — a transformed axis would instead bunch the
labels. Contours are unaffected and stay under `contour_binwidth`, mirroring
ODV's separate Contours tab.

The important property is visible in the two chunks above: DIVAnd ran **once**.
`scale_fill_odv()` operates on the finished ggplot object, so every recolour
after that is instant and you can sweep parameters as fast as you would drag
the ODV sliders.

```{r}
#| label: fig-colour-sweep
#| fig-cap: "Rising nonlinearity at a fixed median. No re-interpolation between panels."
#| fig-height: 8
#| echo: true
for (nl in c(0, 0.3, 0.6, 0.9)) {
  print(p + scale_fill_odv(median = cm_median, nonlinearity = nl) +
          ggtitle(sprintf("nonlinearity = %.1f", nl)))
}
```

The host plot's `zlim`, `palette` and `fill_label` are inherited, so a bare
`scale_fill_odv(median = 0.35)` will not silently reset them. Pass any of
them explicitly to override.

Once you have settled on values, fold them into the original call — the two
arguments below are equivalent to adding the scale afterwards:

```{r}
#| label: colour-mapping-inline
#| eval: false
diva_plot_odv(df, "Temp", time_corr = 20, depth_corr = 15,
              colour_median = 0.35, colour_nonlinearity = 0.6)
```

### Automatic colour mapping {#sec-auto-colour}

Rather than tuning `median` and `nonlinearity` by eye, `odv_auto_map()` derives
both from the interpolated field. It treats the ODV transfer curve as a
two-parameter fit to the data's empirical distribution (histogram
equalisation): the palette midpoint is placed at the field's median, and the
S-curve strength is chosen so the curve tracks the empirical CDF.

```{r}
#| label: auto-colour-fit
#| include: false
grid_temp <- diva_plot_odv(df, "Temp", time_corr = 20, depth_corr = 15,
                           return_data = TRUE)
auto_fit  <- odv_auto_map(grid_temp$Temp)
auto_med  <- round(unname(auto_fit["median"]), 2)
auto_nl   <- round(unname(auto_fit["nonlinearity"]), 2)
```

For this temperature field the fit returns a median of `r auto_med` and a
nonlinearity of `r auto_nl`. You can read those values off directly, or let the
scale fit itself from the plot's own grid with the `"auto"` sentinel --- no
re-interpolation, exactly as in @sec-colour-mapping:

```{r}
#| label: fig-colour-auto
#| fig-cap: "Temperature section with an automatically fitted ODV colour map."
p <- diva_plot_odv(df, "Temp", time_corr = 20, depth_corr = 15)
p + scale_fill_odv(median = "auto", nonlinearity = "auto")
```

`"auto"` works per parameter: fix one and fit the other, e.g.
`scale_fill_odv(median = 0.5, nonlinearity = "auto")`. `diva_plot_odv()` accepts
the same sentinel via `colour_median` / `colour_nonlinearity`. For an
optimiser-free estimate on very large grids, pass `method = "iqr"` to
`odv_auto_map()`.

## Contour control {#sec-contours}

Contours are off by default. They can be set when the plot is built, or added
to a finished plot the same way colour mapping is — and for the same reason,
adding them costs nothing:

```{r}
#| label: contour-breaks
cn_breaks <- c(20, 22, 24, 26)
```

```{r}
#| label: fig-contours-breaks
#| fig-cap: !expr sprintf("Explicit contour levels at %s.", paste(cn_breaks, collapse = ", "))
p + odv_contours(breaks = cn_breaks)
```

`p` here is the plot interpolated back in @sec-colour-mapping. Contours at
`r paste(cn_breaks, collapse = ", ")` °C are drawn onto it without touching
DIVAnd.

Even spacing works too, and the two forms are mutually exclusive — `breaks`
wins if you pass both, with a warning:

```{r}
#| label: fig-contours-binwidth
#| fig-cap: "Even contour spacing at 2 °C intervals, labels suppressed."
p + odv_contours(binwidth = 2, labels = FALSE)
```

`odv_contours()` *replaces* whatever contours are already on the plot rather
than adding a second set, so you can call it repeatedly while tuning without
accumulating layers. To take them off entirely:

```{r}
#| label: fig-contours-removed
#| fig-cap: "Contours stripped; raster, sample points, scales and theme untouched."
p + odv_contours(breaks = cn_breaks) + odv_contours_remove()
```

Removal is by geom class, so contour layers you added by hand come off too.

Contours can also be fixed at build time, which is the right place for them
once you have settled on levels:

```{r}
#| label: contours-inline
#| fig-cap: "Manual Contours"
diva_plot_odv(df, "Temp", time_corr = 20, depth_corr = 15,
              add_contours = TRUE, contour_breaks = c(20, 22, 24, 26))
```

Both routes build their layers with the same internal helper, so a plot's
contours are identical whichever you use.

## Annual climatology (cyclic time)

For a seasonal climatology, fold every observation onto one reference year
(keep month/day, discard the year) and set `cyclic_time = TRUE`. DIVAnd then
treats the time axis as cyclic — December and January share correlation support,
so the January/December seam and the edge relaxation are removed together. The
x-axis is drawn as months.

```{r}
#| label: cyclic-time
#| fig-cap: "Seasonal climatology with a cyclic time axis."
# Fold to a single reference year first (month/day preserved)
folded <- df
folded$Date <- as.Date(format(folded$Date, "2001-%m-%d"))

diva_plot_odv(folded, "Temp",
              time_corr = 30, depth_corr = 20, epsilon2 = 0.3,
              cyclic_time = TRUE)
```

`cyclic_time = TRUE` expects folded input: a `Date` span longer than a year
triggers a warning, since the cyclic wrap would otherwise stitch the earliest
and latest calendar days of the whole record together.

## Masking poorly-constrained nodes {#sec-masking}

ODV blanks nodes the observations don't support. `mask = "support"` masks nodes
whose nearest observation is further than `mask_threshold` correlation lengths
away, measured in the anisotropic metric set by `time_corr` and `depth_corr`.
Because the threshold is in correlation lengths, it means the same thing across
datasets and variables.

```{r}
#| label: fig-mask-support
#| fig-cap: "Temperature section with the default support mask."
diva_plot_odv(df, "Temp", time_corr = 20, depth_corr = 15,
              max_depth = 200, mask = "support")
```

```{r}
#| label: mask-vals
#| include: false

grid_masked <- diva_plot_odv(
  df, "Temp", time_corr = 20, depth_corr = 15, max_depth = 200,
  mask = "support", mask_threshold = 2,
  return_data = TRUE, verbose = FALSE
)

ok_obs       <- !is.na(df$Temp) & !is.na(df$Date)
mask_frac    <- round(100 * mean(grid_masked$mask_metric > 2), 1)
mask_frac_1  <- round(100 * mean(grid_masked$mask_metric > 1), 1)
mask_n_casts <- length(unique(df$Date[ok_obs]))
mask_gap     <- round(as.numeric(median(diff(sort(unique(df$Date[ok_obs]))))), 0)
```

Almost nothing is masked --- `r mask_frac`% of nodes. That is the correct
outcome, not a failure. This record has `r mask_n_casts` casts at a median
spacing of `r mask_gap` days against a 20-day correlation length, so consecutive
casts' envelopes overlap and there is essentially nothing the analysis is
inventing. Tightening to one correlation length masks `r mask_frac_1`%.

To see the mask working, thin the record until it is genuinely sparse:

```{r}
#| label: fig-mask-thinned
#| fig-cap: "The same section from every third cast. White cells are masked."
casts   <- sort(unique(df$Date[!is.na(df$Temp) & !is.na(df$Date)]))
df_thin <- df[df$Date %in% casts[seq(1, length(casts), by = 3)], ]

diva_plot_odv(df_thin, "Temp", time_corr = 20, depth_corr = 15,
              max_depth = 200, mask = "support")
```

```{r}
#| label: mask-thin-vals
#| include: false

grid_thin <- diva_plot_odv(
  df_thin, "Temp", time_corr = 20, depth_corr = 15, max_depth = 200,
  mask = "support", mask_threshold = 2,
  return_data = TRUE, verbose = FALSE
)

thin_frac  <- round(100 * mean(grid_thin$mask_metric > 2), 1)
thin_casts <- length(unique(df_thin$Date[!is.na(df_thin$Temp)]))
thin_gap   <- round(as.numeric(median(diff(sort(
  unique(df_thin$Date[!is.na(df_thin$Temp)]))))), 0)
```

At `r thin_casts` casts and a median gap of `r thin_gap` days, `r thin_frac`% of
nodes are masked, and the white pockets fall between casts and below the deepest
samples --- exactly where the analysis has no information.

### The error-based alternative

`mask = "cpme"` uses DIVAnd's clever poor man's error estimate instead of
geometry. It costs one extra analysis, and its threshold is a dimensionless error
in (0, 1) rather than a distance:

```{r}
#| label: fig-mask-cpme
#| fig-cap: "The thinned section masked by the DIVAnd error field."
diva_plot_odv(df_thin, "Temp", time_corr = 20, depth_corr = 15,
              max_depth = 200, mask = "cpme")
```

Because it looks at the analysis rather than at observation positions, `cpme`
responds to things geometry cannot see --- in particular the interaction between
`epsilon2` and the correlation lengths. Its default of 0.8 was calibrated against
`"support"` at its default.

### Sweeping thresholds without re-running

`return_data = TRUE` adds a `mask_metric` column holding the raw metric, so
choosing a threshold costs no DIVAnd time:

```{r}
#| label: mask-sweep
data.frame(
  threshold = c(0.5, 1, 1.5, 2),
  masked    = sapply(c(0.5, 1, 1.5, 2),
                     function(t) mean(grid_thin$mask_metric > t))
)
```

For `"support"` the metric is exact out to twice the threshold you requested and
`Inf` beyond, so you can loosen the cut by up to a factor of two from a single
run. For `"cpme"` it is exact everywhere. With `cyclic_time = TRUE` the mask
wraps at the year boundary, so it stays continuous where the folded field is.

## Extracting the grid

Set `return_data = TRUE` to get the interpolated grid as a tibble instead of a
plot — useful for further analysis or custom visualisation:

```{r}
#| label: return-data
grid <- diva_plot_odv(df, "Temp", time_corr = 20, depth_corr = 15,
                      return_data = TRUE)
head(grid)
```

## Multi-variable workflows

For projects with many variables, use `diva_variable_config()` to create
reusable parameter sets:

```{r}
#| label: multi-var
#| eval: false
# Load the 12-variable NMP defaults
cfg <- nmp_default_config()

# Plot each variable
for (v in names(cfg)) {
  cv <- cfg[[v]]
  p <- diva_plot_odv(
    df         = my_data,
    var        = cv$var,
    time_corr  = cv$time_corr,
    depth_corr = cv$depth_corr,
    epsilon2   = cv$epsilon2,
    transform  = cv$transform
  )
  print(p)
}
```

## How the metric tensor works

A common pitfall when using DIVAnd is setting the metric tensor (`pmn`) to
all ones. This tells DIVAnd that each grid cell is 1 × 1 in physical units,
regardless of actual spacing — which breaks the correlation length
interpretation and produces vertical-stripe artefacts.

divaodv computes `pmn` correctly from the actual grid spacing:

```
dz = max_depth / (n_depth - 1)     # metres per cell
dt = max_time  / (n_time  - 1)     # years per cell
pm = 1 / dz                        # points per metre
pn = 1 / dt                        # points per year
```

This means correlation lengths are always honoured regardless of grid
resolution — you can change `time_resolution` from 365 to 52 without
affecting the physical smoothing behaviour.
